---
jupyter:
  jupytext:
    text_representation:
      extension: .Rmd
      format_name: rmarkdown
      format_version: '1.2'
      jupytext_version: 1.19.1
  kernelspec:
    display_name: R
    language: R
    name: ir
---

# Prior Sensitivity in Bayesian Logistic Regression


## 1. Data & EDA

```{r}
library(MASS)        # Built-in dataset — provides the birthwt dataset
library(rstanarm)    # Bayesian logistic regression via Stan
library(bayesplot)   # Visualizing posterior distributions and predictive checks
library(pROC)        # Computing and comparing ROC curves and AUC
library(ggplot2)     # Custom data visualization
library(dplyr)       # Data manipulation for subsampling experiment
library(tidyr)
data(birthwt)
```

### Dataset Overview

```{r}
str(birthwt)
```

```{r}
birthwt$race  <- as.factor(birthwt$race)
birthwt$smoke <- as.factor(birthwt$smoke)
birthwt$ht    <- as.factor(birthwt$ht)
birthwt$ui    <- as.factor(birthwt$ui)

summary(birthwt)
```

```{r}
sum(is.na(birthwt))

table(birthwt$low)
```

### Feature Description


#### Class balance

```{r}
ggplot(data = birthwt, aes(x = factor(low, labels = c("Normal", "Low Birth Weight")))) +
  geom_bar(fill = c("#4CAF50", "#F44336")) +
  geom_text(stat = "count", aes(label = after_stat(count)), vjust = -0.5) +
  labs(title = "Class Balance of Low Birth Weight",
       x = "Birth Weight Status",
       y = "Count") +
  theme_minimal()
```

#### Continuous features

```{r}
features <- c("age", "lwt", "ptl", "ftv")

df.long <- pivot_longer(birthwt[, c("low", features)],
                        cols = -low,
                        names_to = "feature",
                        values_to = "value")

ggplot(data = df.long, aes(x = factor(low), y = value, fill = factor(low))) +
  geom_boxplot(outlier.size = 0.5, alpha = 0.7) +
  facet_wrap(~ feature, scales = "free_y", ncol = 2) +
  scale_fill_manual(values = c("0" = "#4CAF50", "1" = "#F44336")) +
  labs(title = "Continuous Feature Distribution by Low Birth Weight",
       x = "Low Birth Weight (0 = No, 1 = Yes)", 
       y = "Value") +
  theme_minimal() +
  theme(legend.position = "none")
```

#### Categorical features

```{r}
features.cat <- c("race", "smoke", "ht", "ui")

df.cat <- pivot_longer(birthwt[, c("low", features.cat)],
                       cols = -low,
                       names_to = "feature",
                       values_to = "value")

ggplot(data = df.cat, aes(x = factor(value), fill = factor(low))) +
  geom_bar(position = "dodge") +
  facet_wrap(~ feature, scales = "free_x", ncol = 2) +
  scale_fill_manual(values = c("0" = "#4CAF50", "1" = "#F44336"),
                    labels = c("Normal", "Low Birth Weight")) +
  labs(title = "Categorical Feature Distribution by Low Birth Weight",
       x = "Value",
       y = "Count",
       fill = "Birth Weight") +
  theme_minimal()
```

#### Feature Correlation

```{r}
features.only <- birthwt[, c("age", "lwt", "ptl", "ftv")]
corr.matrix   <- cor(features.only)

corrplot::corrplot(corr.matrix,
                   method     = "color",
                   type       = "upper",
                   tl.cex     = 0.7,
                   tl.col     = "black",
                   addCoef.col = "black",
                   number.cex = 0.8,
                   title      = "Correlation Matrix of Continuous Features",
                   mar        = c(0, 0, 1, 0))
```

## 2. Helper Functions

```{r}
run_stan_glm <- function(data, prior, prior_int) {
  stan_glm(
    low ~ age + lwt + race + smoke + ptl + ht + ui + ftv,
    data            = data,
    family          = binomial(link = "logit"),
    prior           = prior,
    prior_intercept = prior_int,
    seed            = 42,
    refresh         = 0
  )
}
```

```{r}
get_posterior_summary <- function(fit, model_name) {
  ps <- as.data.frame(fit$stan_summary)
  ps <- ps[!rownames(ps) %in% c("mean_PPD", "log-posterior"), ]
  data.frame(
    Model    = model_name,
    Variable = rownames(ps),
    Mean     = round(ps$mean, 3),
    CI_2.5   = round(ps$`2.5%`, 3),
    CI_97.5  = round(ps$`97.5%`, 3),
    OR       = round(exp(ps$mean), 3)
  )
}
```

```{r}
get_auc <- function(fit, data) {
  pred      <- colMeans(posterior_epred(fit))
  roc.obj   <- roc(data$low, pred, quiet = TRUE)
  as.numeric(auc(roc.obj))
}
```

```{r}
loo_safe <- function(fit) {
  tryCatch({
    l      <- suppressWarnings(loo(fit))
    looic  <- l$estimates["looic", "Estimate"]
    k_warn <- sum(l$diagnostics$pareto_k > 0.7)
    list(looic = round(looic, 3), k_warn = k_warn)
  }, error = function(e) list(looic = NA, k_warn = NA))
}
```

# Analysis 1: Prior Sensitivity (Full Data, n = 189)


## Prior Specification


### Diffuse Prior — N(0, 100)

```{r}
prior.diffuse <- normal(0, 100)
prior_int.diffuse <- normal(0, 100)
fit.diffuse <- run_stan_glm(birthwt, prior.diffuse, prior_int.diffuse)
```

### Weakly Informative Prior — N(0, 2.5)

```{r}
prior.weak        <- normal(0, 2.5)
prior_int.weak    <- normal(0, 10)
fit.weak          <- run_stan_glm(birthwt, prior.weak, prior_int.weak)
```

### Clinically Informative Prior

```{r}
prior.informative <- normal(
  location = c(0,    # age
               0,    # lwt
               0,    # race2
               0,    # race3
               1,    # smoke1
               0,    # ptl
               1,    # ht1
               0,    # ui1
               0),   # ftv
  scale    = c(1,    # age
               1,    # lwt
               1,    # race2
               1,    # race3
               0.5,  # smoke1
               1,    # ptl
               0.5,  # ht1
               1,    # ui1
               1)    # ftv
)
prior_int.informative <- normal(0, 5)
fit.informative       <- run_stan_glm(birthwt, prior.informative, prior_int.informative)
```

## MCMC Diagnostics


### Trace Plots


#### Diffuse Prior — N(0, 100)

```{r}
mcmc_trace(as.array(fit.diffuse),
           pars = c("age", "lwt", "smoke1", "ht1")) +
  ggtitle("Trace Plot - Diffuse Prior")
```

#### Weakly Informative Prior — N(0, 2.5)

```{r}
mcmc_trace(as.array(fit.weak),
           pars = c("age", "lwt", "smoke1", "ht1")) +
  ggtitle("Trace Plot - Weakly Informative Prior")
```

#### Clinically Informative Prior

```{r}
mcmc_trace(as.array(fit.informative),
           pars = c("age", "lwt", "smoke1", "ht1")) +
  ggtitle("Trace Plot - Clinically Informative Prior")
```

### Rhat Convergence Summary

```{r}
result.rhat <- data.frame(
  Variable    = rownames(fit.diffuse$stan_summary),
  Rhat_Diffuse     = round(fit.diffuse$stan_summary[, "Rhat"], 3),
  Rhat_Weak        = round(fit.weak$stan_summary[, "Rhat"], 3),
  Rhat_Informative = round(fit.informative$stan_summary[, "Rhat"], 3)
)
print(result.rhat, row.names = FALSE)
```

## Posterior Summary


### Coefficient Estimates

```{r}
print(fit.diffuse, digits = 3)
```

```{r}
print(fit.weak, digits = 3)
```

```{r}
print(fit.informative, digits = 3)
```

### Coefficient Estimates Across Priors

```{r}
result.coef.compare <- data.frame(
  Variable    = names(coef(fit.diffuse)),
  Diffuse     = round(coef(fit.diffuse), 3),
  Weak        = round(coef(fit.weak), 3),
  Informative = round(coef(fit.informative), 3)
)
print(result.coef.compare, row.names = FALSE)
```

```{r}
result.posterior <- rbind(
  get_posterior_summary(fit.diffuse,     "Diffuse"),
  get_posterior_summary(fit.weak,        "Weak"),
  get_posterior_summary(fit.informative, "Informative")
)
print(result.posterior, row.names = FALSE)
```

### Coefficient Plot  


#### Continuous Predictors

```{r}
df.plot <- result.posterior[result.posterior$Variable != "(Intercept)", ]
df.plot$Model <- factor(df.plot$Model, levels = c("Diffuse", "Weak", "Informative"))

vars.continuous <- c("age", "lwt", "ftv")
df.continuous <- df.plot[df.plot$Variable %in% vars.continuous, ]

ggplot(df.continuous, aes(x = Mean, y = Variable, color = Model, shape = Model)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  geom_pointrange(aes(xmin = CI_2.5, xmax = CI_97.5),
                  position = position_dodge(width = 0.5),
                  size = 0.5) +
  scale_color_manual(values = c("Diffuse"     = "#E41A1C",
                                "Weak"        = "#377EB8",
                                "Informative" = "#4DAF4A")) +
  scale_shape_manual(values = c("Diffuse" = 16, "Weak" = 17, "Informative" = 15)) +
  labs(title    = "Posterior Coefficients Across Priors (n = 189) — Continuous Predictors",
       subtitle = "Point = posterior mean, bars = 95% credible interval",
       x = "Coefficient (log-odds)", y = NULL,
       color = "Prior", shape = "Prior") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

#### Categorical / Binary Predictors

```{r}
vars.binary <- c("race2", "race3", "smoke1", "ptl", "ht1", "ui1")
df.binary <- df.plot[df.plot$Variable %in% vars.binary, ]

ggplot(df.binary, aes(x = Mean, y = Variable, color = Model, shape = Model)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  geom_pointrange(aes(xmin = CI_2.5, xmax = CI_97.5),
                  position = position_dodge(width = 0.5),
                  size = 0.5) +
  scale_color_manual(values = c("Diffuse"     = "#E41A1C",
                                "Weak"        = "#377EB8",
                                "Informative" = "#4DAF4A")) +
  scale_shape_manual(values = c("Diffuse" = 16, "Weak" = 17, "Informative" = 15)) +
  labs(title    = "Posterior Coefficients Across Priors (n = 189) — Categorical Predictors",
       subtitle = "Point = posterior mean, bars = 95% credible interval",
       x = "Coefficient (log-odds)", y = NULL,
       color = "Prior", shape = "Prior") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

### LOO Model Comparison

```{r}
loo.diffuse     <- loo(fit.diffuse)
loo.weak        <- loo(fit.weak)
loo.informative <- loo(fit.informative)

loo.compare <- loo_compare(loo.diffuse, loo.weak, loo.informative)

result.loo <- data.frame(
  Model    = rownames(loo.compare),
  elpd_loo = round(loo.compare[, "elpd_loo"], 2),
  p_loo    = round(loo.compare[, "p_loo"], 2),
  looic    = round(loo.compare[, "looic"], 2)
)
print(result.loo, row.names = FALSE)
```

### AUC Comparison

```{r}
result.auc <- data.frame(
  Model = c("Diffuse", "Weak", "Informative"),
  AUC   = round(c(
    get_auc(fit.diffuse,     birthwt),
    get_auc(fit.weak,        birthwt),
    get_auc(fit.informative, birthwt)
  ), 3)
)
print(result.auc, row.names = FALSE)
```

# Analysis 2: Subsampling Experiment


## Fit Models Across Sample Sizes

```{r}
sample.sizes <- c(20, 40, 80, 189)
results.sub  <- list()
auc.list     <- list()

for (n in sample.sizes) {
  set.seed(42)
  data.sub <- if (n < 189) birthwt[sample(nrow(birthwt), n), ] else birthwt

  fit.sub.diffuse     <- run_stan_glm(data.sub, prior.diffuse,     prior_int.diffuse)
  fit.sub.weak        <- run_stan_glm(data.sub, prior.weak,        prior_int.weak)
  fit.sub.informative <- run_stan_glm(data.sub, prior.informative, prior_int.informative)

  # Analysis 2：posterior summary
  results.sub[[as.character(n)]] <- rbind(
    get_posterior_summary(fit.sub.diffuse,     "Diffuse"),
    get_posterior_summary(fit.sub.weak,        "Weak"),
    get_posterior_summary(fit.sub.informative, "Informative")
  )

  # Analysis 3：AUC + LOO
  auc.list[[as.character(n)]] <- data.frame(
    n      = n,
    Model  = c("Diffuse", "Weak", "Informative"),
    AUC    = round(c(
      get_auc(fit.sub.diffuse,     data.sub),
      get_auc(fit.sub.weak,        data.sub),
      get_auc(fit.sub.informative, data.sub)
    ), 3),
    LOOIC  = c(
      loo_safe(fit.sub.diffuse)$looic,
      loo_safe(fit.sub.weak)$looic,
      loo_safe(fit.sub.informative)$looic
    ),
    k_warn = c(
      loo_safe(fit.sub.diffuse)$k_warn,
      loo_safe(fit.sub.weak)$k_warn,
      loo_safe(fit.sub.informative)$k_warn
    )
  )
  # Save models for app
  saveRDS(fit.sub.diffuse,     paste0("../models/fit_diffuse_n",     n, ".rds"))
  saveRDS(fit.sub.weak,        paste0("../models/fit_weak_n",        n, ".rds"))
  saveRDS(fit.sub.informative, paste0("../models/fit_informative_n", n, ".rds"))

  cat("Done: n =", n, "\n")
}

df.sub <- do.call(rbind, lapply(names(results.sub), function(n) {
  df <- results.sub[[n]]; df$n <- as.integer(n); df
}))

df.auc <- do.call(rbind, auc.list)
```

## Coefficient Convergence: smoke1 & ui1 & ht1

```{r}
print(df.sub[df.sub$Variable %in% c("smoke1", "ui1", "ht1"), 
             c("n", "Model", "Variable", "Mean", "CI_2.5", "CI_97.5")], 
      row.names = FALSE)
```

```{r}
df.smoke <- df.sub[df.sub$Variable == "smoke1", ]
df.smoke$Model <- factor(df.smoke$Model, levels = c("Diffuse", "Weak", "Informative"))
df.smoke$n     <- factor(df.smoke$n, levels = c(20, 40, 80, 189))

ggplot(df.smoke, aes(x = n, y = Mean, color = Model, group = Model)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_ribbon(aes(ymin = CI_2.5, ymax = CI_97.5, fill = Model),
              alpha = 0.15, color = NA) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2.5) +
  scale_color_manual(values = c("Diffuse"     = "#E41A1C",
                                "Weak"        = "#377EB8",
                                "Informative" = "#4DAF4A")) +
  scale_fill_manual(values  = c("Diffuse"     = "#E41A1C",
                                "Weak"        = "#377EB8",
                                "Informative" = "#4DAF4A")) +
  labs(title    = "Coefficient Convergence — Smoking (smoke1)",
       subtitle = "Shaded band = 95% credible interval",
       x = "Sample Size (n)", y = "Coefficient (log-odds)",
       color = "Prior", fill = "Prior") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

```{r}
df.ui <- df.sub[df.sub$Variable == "ui1", ]
df.ui$Model <- factor(df.ui$Model, levels = c("Diffuse", "Weak", "Informative"))
df.ui$n     <- factor(df.ui$n, levels = c(20, 40, 80, 189))

ggplot(df.ui, aes(x = n, y = Mean, color = Model, group = Model)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_ribbon(aes(ymin = CI_2.5, ymax = CI_97.5, fill = Model),
              alpha = 0.15, color = NA) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2.5) +
  scale_color_manual(values = c("Diffuse"     = "#E41A1C",
                                "Weak"        = "#377EB8",
                                "Informative" = "#4DAF4A")) +
  scale_fill_manual(values  = c("Diffuse"     = "#E41A1C",
                                "Weak"        = "#377EB8",
                                "Informative" = "#4DAF4A")) +
  labs(title    = "Coefficient Convergence — Uterine Irritability (ui1)",
       subtitle = "Shaded band = 95% credible interval",
       x = "Sample Size (n)", y = "Coefficient (log-odds)",
       color = "Prior", fill = "Prior") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

```{r}
df.ht <- df.sub[df.sub$Variable == "ht1", ]
df.ht$Model <- factor(df.ht$Model, levels = c("Diffuse", "Weak", "Informative"))
df.ht$n     <- factor(df.ht$n, levels = c(20, 40, 80, 189))

ggplot(df.ht, aes(x = n, y = Mean, color = Model, group = Model)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_ribbon(aes(ymin = CI_2.5, ymax = CI_97.5, fill = Model),
              alpha = 0.15, color = NA) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2.5) +
  scale_color_manual(values = c("Diffuse"     = "#E41A1C",
                                "Weak"        = "#377EB8",
                                "Informative" = "#4DAF4A")) +
  scale_fill_manual(values  = c("Diffuse"     = "#E41A1C",
                                "Weak"        = "#377EB8",
                                "Informative" = "#4DAF4A")) +
  labs(title    = "Coefficient Convergence — Hypertension (ht1)",
       subtitle = "Shaded band = 95% credible interval",
       x = "Sample Size (n)", y = "Coefficient (log-odds)",
       color = "Prior", fill = "Prior") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

## Robustness Check: Repeated Subsampling

```{r}
R             <- 100
sub.sizes.rep <- c(20, 40, 80)
track.vars    <- c("smoke1", "ui1", "ht1")
explode.cut   <- 10

build_formula <- function(data) {
  terms_all <- c("age","lwt","race","smoke","ptl","ht","ui","ftv")
  keep <- sapply(terms_all, function(v) {
    x <- data[[v]]
    if (is.factor(x)) nlevels(droplevels(x)) >= 2 else length(unique(x)) >= 2
  })
  reformulate(terms_all[keep], response = "low")
}

make_diffuse <- function(form, data) {
  nm <- colnames(model.matrix(form, data))
  nm <- nm[nm != "(Intercept)"]
  normal(location = rep(0, length(nm)), scale = rep(10, length(nm)))
}

make_weak <- function(form, data) {
  nm <- colnames(model.matrix(form, data))
  nm <- nm[nm != "(Intercept)"]
  normal(location = rep(0, length(nm)), scale = rep(2.5, length(nm)))
}

make_informative <- function(form, data) {
  nm <- colnames(model.matrix(form, data))
  nm <- nm[nm != "(Intercept)"]
  normal(location = ifelse(nm %in% c("smoke1","ht1"), 1, 0),
         scale    = ifelse(nm %in% c("smoke1","ht1"), 0.5, 1))
}
prior_int.shared <- normal(0, 10)

fit_one <- function(data, form, prior, prior_int) {
  stan_glm(form, data = data, family = binomial("logit"),
           prior = prior, prior_intercept = prior_int,
           seed = 42, refresh = 0)
}

rep.rows <- list(); k <- 0

for (n in sub.sizes.rep) {
  for (r in seq_len(R)) {
    set.seed(n * 1000 + r)
    data.sub <- droplevels(birthwt[sample(nrow(birthwt), n), , drop = FALSE])
    form     <- build_formula(data.sub)

    fits <- list(
      Diffuse     = tryCatch(
        fit_one(data.sub, form, make_diffuse(form, data.sub),     prior_int.shared),
        error = function(e) NULL),
      Weak        = tryCatch(
        fit_one(data.sub, form, make_weak(form, data.sub),        prior_int.shared),
        error = function(e) NULL),
      Informative = tryCatch(
        fit_one(data.sub, form, make_informative(form, data.sub), prior_int.shared),
        error = function(e) NULL)
    )

    for (mdl in names(fits)) {
      for (v in track.vars) {
        k   <- k + 1
        fit <- fits[[mdl]]
        defined <- !is.null(fit) && (v %in% names(coef(fit)))
        m   <- if (defined) as.numeric(coef(fit)[v]) else NA_real_
        rep.rows[[k]] <- data.frame(
          n        = n,
          rep      = r,
          Model    = mdl,
          Variable = v,
          Mean     = m,
          defined  = defined,
          exploded = isTRUE(abs(m) > explode.cut)
        )
      }
    }
  }
  cat("Done repeated subsampling: n =", n, "\n")
}

df.rep <- do.call(rbind, rep.rows)
saveRDS(df.rep, "../models/df_rep.rds")
```

```{r}
df.rep %>%
  group_by(n, Model, Variable) %>%
  summarise(
    pct_undef   = round(100 * mean(!defined), 1),
    pct_explode = round(100 * mean(exploded[defined], na.rm = TRUE), 1),  # ← 只看 defined
    median_mean = round(median(Mean[defined & !exploded], na.rm = TRUE), 2),
    .groups = "drop"
  ) %>%
  filter(Model == "Diffuse") %>% arrange(Variable, n) %>% print(n = Inf)
```

```{r}
df.rep %>%
  filter(Model == "Diffuse") %>%
  group_by(n, Variable) %>%
  summarise(
    n_explode = sum(exploded[defined], na.rm = TRUE),
    .groups = "drop"
  ) %>%
  ggplot(aes(factor(n), n_explode, color = Variable, group = Variable)) +
  geom_line(linewidth = 0.9) + geom_point(size = 3) +
  labs(title = "Diffuse-prior coefficient instability across 100 subsamples",
     x = "Sample size (n)", 
     y = "Count of |β| > 10 (of 100 draws)", 
     color = "Predictor") +
  theme_minimal(base_size = 12) + theme(legend.position = "bottom")
```

# Analysis 3: Predictive Performance Across Sample Sizes


## AUC by Sample Size and Prior

```{r}
print(df.auc, row.names = FALSE)
```

```{r}
df.auc %>%
  ggplot(aes(factor(n), LOOIC, color = Model, group = Model)) +
  geom_line(linewidth = 0.9) + geom_point(size = 3) +
  labs(title = "LOO-CV by Sample Size: Informative Prior Wins",
       x = "Sample size (n)", y = "LOOIC (lower = better)") +
  theme_minimal(base_size = 12) + theme(legend.position = "bottom")
```

```{r}
df.auc %>%
  ggplot(aes(factor(n), k_warn, color = Model, group = Model)) +
  geom_line(linewidth = 0.9) + geom_point(size = 3) +
  labs(title = "Pareto k Diagnostics: Diffuse Prior Unreliable at Small n",
       x = "Sample size (n)", y = "Count (Pareto k > 0.7)") +
  theme_minimal(base_size = 12) + theme(legend.position = "bottom")
```

```{r}
df.auc$Model <- factor(df.auc$Model, levels = c("Diffuse", "Weak", "Informative"))
df.auc$n     <- factor(df.auc$n, levels = c(20, 40, 80, 189))

ggplot(df.auc, aes(x = n, y = AUC, color = Model, group = Model)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 3) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "grey50") +
  scale_color_manual(values = c("Diffuse"     = "#E41A1C",
                                "Weak"        = "#377EB8",
                                "Informative" = "#4DAF4A")) +
  scale_y_continuous(limits = c(0.4, 1.0), breaks = seq(0.4, 1.0, 0.1)) +
  labs(title    = "AUC by Sample Size and Prior",
       subtitle = "In-sample AUC; dashed line = random classifier",
       x = "Sample Size (n)", y = "AUC",
       color = "Prior") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

# Export Models

```{r}
saveRDS(fit.diffuse,     "../models/fit_diffuse.rds")
saveRDS(fit.weak,        "../models/fit_weak.rds")
saveRDS(fit.informative, "../models/fit_informative.rds")
cat("Models saved.\n")
```
